---
title: "Biostat 203B Homework 2"
subtitle: Due Feb 10 @ 11:59PM
author: Sylvia Wang, 105118268
format:
  html:
    theme: cosmo
    number-sections: true
    toc: true
    toc-depth: 4
    toc-location: left
    code-fold: false
knitr:
  opts_chunk: 
    cache: false    
    echo: true
    fig.align: 'center'
    fig.width: 6
    fig.height: 4
    message: FALSE
---

Display machine information for reproducibility:
```{r}
sessionInfo()
```

Load necessary libraries (you can add more as needed).
```{r setup}
library(data.table)
library(lubridate)
library(R.utils)
library(tidyverse)
```

MIMIC data location
```{r}
mimic_path <- "~/mimic"
```

In this exercise, we use tidyverse (ggplot2, dplyr, etc) to explore the [MIMIC-IV](https://mimic.mit.edu/docs/iv/) data introduced in [homework 1](https://ucla-biostat-203b.github.io/2023winter/hw/hw1/hw1.html) and to build a cohort of ICU stays.

Display the contents of MIMIC data folder. 
```{r}
system(str_c("ls -l ", mimic_path, "/"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/core"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/hosp"), intern = TRUE)
system(str_c("ls -l ", mimic_path, "/icu"), intern = TRUE)
```

## Q1. `read.csv` (base R) vs `read_csv` (tidyverse) vs `fread` (data.table)

There are quite a few utilities in R for reading plain text data files. Let us test the speed of reading a moderate sized compressed csv file, `admissions.csv.gz`, by three programs: `read.csv` in base R, `read_csv` in tidyverse, and `fread` in the popular data.table package. 

Which function is fastest? Is there difference in the (default) parsed data types? (Hint: R function `system.time` measures run times.)

For later questions, we stick to the `read_csv` in tidyverse.

```{r}
system.time(ad1 <- read.csv("~/mimic/core/admissions.csv.gz"))
```

```{r}
system.time(ad2 <- read_csv("~/mimic/core/admissions.csv.gz"))
```

```{r}
system.time(ad3 <- fread("~/mimic/core/admissions.csv.gz"))
```

`fread` is the fastest function.

(default) parsed data types:
```{r}
str(ad1)
```

```{r}
str(ad2)
```

```{r}
str(ad3)
```
When using `read.csv`, the file is read in as a data frame. The data formats of the columns within the data frame are integer (int) (subject_id, hadm_id, hospital_expire_flag) and character (chr) (admittime, dischtime, deathtime, admission_type, admission_location, discharge_location, insurance, language, marital_status, ethnicity, edregtime, edouttime).

When using `read_csv`, the file's read-in format can be regarded as spec_tbl_df, tbl_df, tbl (tibble), or data frame. The data format of the columns within the tibble are called numeric (num), character (chr), and POSIXct. POSIXct is the data format for date and time (admittime, dischtime, deathtime, edregtime, edouttime). It stores date and time in seconds with the number of seconds beginning at 1 January 1970.

When using `fread`, the file's read-in format can be regarded as data table or data frame. The data formats of the columns are integer (int), character (chr), and POSIXct.

The numeric and integer formats are interchangeable for the columns with pure number values. Date and time values are either regarded as character when using `read.csv`, or recorded in the POSIXct format specific for date and time when the file is being imported using `read_csv` or `fread`. Columns with normal texts are always in the character format regardless of importing method.

## Q2. ICU stays

`icustays.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/icustays/>) contains data about Intensive Care Units (ICU) stays. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/icustays.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `icustays.csv.gz` as a tibble `icustays_tble`. 
```{r}
icustays_tble <- read_csv("~/mimic/icu/icustays.csv.gz")
```

2. How many unique `subject_id`? Can a `subject_id` have multiple ICU stays? 
```{r}
length(unique(icustays_tble$subject_id))
```
There are 53150 unique `subject_id`.

```{r}
head(table(icustays_tble$subject_id),10)
```
Yes, a `subject_id` can have multiple ICU stays. As demonstrated by the first 10 ids here, some of them have more than 1 ICU stays.

3. Summarize the number of ICU stays per `subject_id` by graphs. 

```{r}
stays <- icustays_tble %>%
  group_by(subject_id) %>%
  summarize(stay_count = n()) 

ggplot(data = stays) + 
  geom_bar(mapping = aes(x = stay_count)) +
  xlab("Number of ICU stays") +
  ylab("Number of subjects \n staying the certain number of ICU stays") 
```
Since the majority of subjects only had one ICU stay, it is hard to see those with more stay on the graph.

I thus also tried showing the actual numbers of subjects having certain number of ICU stays on top of each bar in the histogram, but the result is not very pretty either (but at least we can see those having a large number of stays):
```{r}
count <- stays %>% 
  group_by(stay_count) %>%
  summarize(subject_count = n())

ggplot(count, aes(x = stay_count, y = subject_count)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = subject_count, vjust = -0.25)) +
  xlab("Number of ICU stays") +
  ylab("Number of subjects \n staying the certain number of ICU stays")
```

4. For each `subject_id`, let's only keep the first ICU stay in the tibble `icustays_tble`. (Hint: `slice_min` and `slice_max` may take long. Think alternative ways to achieve the same function.)

```{r}
icustays_tble <- icustays_tble %>% 
  group_by(subject_id) %>% 
  arrange(subject_id, intime) %>% 
  slice(1)
icustays_tble
```

## Q3. `admission` data

Information of the patients admitted into hospital is available in `admissions.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/admissions/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/admissions.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `admissions.csv.gz` as a tibble `admissions_tble`.
```{r}
admissions_tble <- read_csv("~/mimic/core/admissions.csv.gz")
```

2. Let's only keep the admissions that have a match in `icustays_tble` according to `subject_id` and `hadmi_id`.

```{r}

```


3. Summarize the following variables by graphics. 

    - admission year  
    - admission month  
    - admission month day  
    - admission week day  
    - admission hour (anything unusual?)  
    - admission minute (anything unusual?)  
    - length of hospital stay (anything unusual?)    
    
## Q4. `patients` data

Patient information is available in `patients.csv.gz`. See <https://mimic.mit.edu/docs/iv/modules/core/patients/> for details of each field in this file. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/core/patients.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Import `patients.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/core/patients/>) as a tibble `patients_tble` and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

2. Summarize variables `gender` and `anchor_age`, and explain any patterns you see.

## Q5. Lab results

`labevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/hosp/labevents/>) contains all laboratory measurements for patients. The first 10 lines are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/labevents.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```
`d_labitems.csv.gz` is the dictionary of lab measurements. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/hosp/d_labitems.csv.gz"), 
    " | head"
    ), 
  intern = TRUE
)
```

1. Find how many rows are in `labevents.csv.gz`.

2. We are interested in the lab measurements of creatinine (50912), potassium (50971), sodium (50983), chloride (50902), bicarbonate (50882), hematocrit (51221), white blood cell count (51301), and glucose (50931). Retrieve a subset of `labevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `labevents_tble`. 

    Hint: `labevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer 5-10 minutes to ingest this file. If your computer really has trouble importing `labevents.csv.gz`, you can import from the reduced data file `labevents_filtered_itemid.csv.gz`.

3. Further restrict `labevents_tble` to the first lab measurement during the ICU stay. 

4. Summarize the lab measurements by appropriate numerics and graphics. 

## Q6. Vitals from charted events

`chartevents.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/chartevents/>) contains all the charted data available for a patient. During their ICU stay, the primary repository of a patient’s information is their electronic chart. The `itemid` variable indicates a single measurement type in the database. The `value` variable is the value measured for `itemid`. The first 10 lines of `chartevents.csv.gz` are
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/chartevents.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```
`d_items.csv.gz` (<https://mimic.mit.edu/docs/iv/modules/icu/d_items/>) is the dictionary for the `itemid` in `chartevents.csv.gz`. 
```{r}
system(
  str_c(
    "zcat < ", 
    str_c(mimic_path, "/icu/d_items.csv.gz"), 
    " | head"), 
  intern = TRUE
)
```

1. We are interested in the vitals for ICU patients: heart rate (220045), mean non-invasive blood pressure (220181), systolic non-invasive blood pressure (220179), body temperature in Fahrenheit (223761), and respiratory rate (220210). Retrieve a subset of `chartevents.csv.gz` only containing these items for the patients in `icustays_tble` as a tibble `chartevents_tble`.

    Hint: `chartevents.csv.gz` is a data file too big to be read in by the `read_csv` function in its default setting. Utilize the `col_select` option in the `read_csv` function to reduce the memory burden. It took my computer >15 minutes to ingest this file. If your computer really has trouble importing `chartevents.csv.gz`, you can import from the reduced data file `chartevents_filtered_itemid.csv.gz`.

2. Further restrict `chartevents_tble` to the first vital measurement during the ICU stay. 

3. Summarize these vital measurements by appropriate numerics and graphics. 

## Q7. Putting things together

Let us create a tibble `mimic_icu_cohort` for all ICU stays, where rows are the first ICU stay of each unique adult (age at admission > 18) and columns contain at least following variables  

- all variables in `icustays.csv.gz`  
- all variables in `admission.csv.gz`  
- all variables in `patients.csv.gz`  
- first lab measurements during ICU stay  
- first vital measurements during ICU stay
- an indicator variable `thirty_day_mort` whether the patient died within 30 days of hospital admission (30 day mortality)

## Q8. Exploratory data analysis (EDA)

Summarize following information using appropriate numerics or graphs.

- `thirty_day_mort` vs demographic variables (ethnicity, language, insurance, marital_status, gender, age at hospital admission)

- `thirty_day_mort` vs first lab measurements

- `thirty_day_mort` vs first vital measurements

- `thirty_day_mort` vs first ICU unit

